### 2015-06-01 papers/histogram/thesis/thesis.tex
trimming a couple dependencies for thesis.tex
shortened abstract for the honors college; final version of thesis

%%% figures
\usepackage{graphicx,float,subcaption} % floats, etc.
\usepackage{footnote} % for footnotes in floats
\usepackage[font=small,labelfont=bf]{caption} % caption text options

@@

\renewcommand{\title}{Optimizing Monte Carlo simulation of the
  square-well fluid\xspace}
\renewcommand{\author}{Michael A. Perlin\xspace}
\newcommand{\degree}{Honors Baccalaureate of Science
  in Physics\xspace}
\renewcommand{\date}{02 June 2015\xspace}
\newcommand{\email}{\href{mailto:mika.perlin@gmail.com}{mika.perlin@gmail.com}\xspace}

@@

\author for the degree of \underline{\degree} presented on
\date. Title: \underline{\title}.

@@

We identify and develop efficient numerical methods for determining
thermodynamic properties of the square-well fluid in order to test
square-well density functional theories. The liquid-vapor phase
interface is an interesting regime for testing density functional
theories, but unbiased Monte Carlo simulations are incapable of
effectively sampling the low energy fluid states which dominate the
partition function at low liquid-state temperatures. Previous works
have developed several generic Monte Carlo histogram methods for
collecting statistics on low energy system states, but little or no
literature exists on these methods' systematic comparison, as well as
their application to the square-well fluid. The square-well fluid in
particular introduces application challenges not manifest in
traditional models for testing and benchmarking such numerical
techniques (e.g. the Ising model).

@@

We implement our own ``simple flat'' method, the Wang-Landau method,
transition matrix Monte Carlo, and a modified version of the optimized
ensemble. Performance of each method is determined by low energy
sampling rates and maximum errors in several computed system
properties. We find that Wang-Landau potentially performs better than
other histogram methods, but fails catastrophically without a
predetermined energy range. The simple flat method and transition
matrix Monte Carlo give results comparable to successful Wang-Landau
simulations, but simple flat has some anomalous cases with large
errors. Finally, our implementation of the optimized ensemble using a
transition matrix results in worse performance over the
straightforward transition matrix Monte Carlo method.

@@

Key words: Monte Carlo methods, Square-well fluid, histogram methods

@@

\underline{\degree} project of \author presented on \date.

@@

\hrule ~\\
David Roundy, Mentor, representing the Department of Physics

@@

\hrule ~\\
Henri J. F. Jansen, Committee Member, representing the Department of
Physics

@@

\hrule ~\\
David H. McIntyre, Committee Member, representing the Department of
Physics

@@

Algorithm \ref{alg:wang_landau} spells out the Wang-Landau
method. This algorithm includes four free parameters which affect
initialization process: the number of iterations $n$ to run in step
\ref{alg:wang_landau_sim}, the initial factor $f_0$, the update factor
$u$, and the flatness condition $C\sp{H}$. The cutoff $c$ determines
the condition for Algorithm \ref{alg:wang_landau} to terminate.

>>>>

%%% figures
\usepackage{graphicx,float} % floats, etc.
\usepackage[font=small,labelfont=bf]{caption} % caption text options

@@

\renewcommand{\title}{Optimizing Monte Carlo simulation of the
  square-well fluid\xspace}
\renewcommand{\author}{Michael A. Perlin\xspace}
\newcommand{\degree}{Honors Baccalaureate of Science
  in Physics\xspace}
\renewcommand{\date}{02 June 2015\xspace}
\newcommand{\email}{\href{mailto:mika.perlin@gmail.com}
  {mika.perlin@gmail.com}\xspace}

@@

\underline{\author} for the degree of \underline{\degree} presented on
\underline{\date}. Title: \underline{\title}.

@@

We identify and develop efficient Monte Carlo methods for determining
thermodynamic properties of the square-well fluid in order to test
square-well density functional theories near the critical
point. Previous works have developed generic so-called histogram
methods for collecting statistics on low energy system states, but
little or no literature exists on these methods' systematic
comparison, as well as their application to the square-well fluid. The
square-well fluid in particular introduces application challenges not
manifest in traditional models for testing and benchmarking such
numerical techniques (e.g. the Ising model).

@@

We implement our own ``simple flat'' method, the Wang-Landau method,
transition matrix Monte Carlo (TMMC), and a modified version of the
optimized ensemble. The performance of each method is measured by low
energy sampling rates and maximum errors in computed system
properties. We find that Wang-Landau potentially performs better than
other histogram methods, but fails catastrophically without a
predetermined energy range. The simple flat method and TMMC give
results comparable to successful Wang-Landau simulations, but simple
flat has some anomalous cases with large errors. Finally, our
implementation of the optimized ensemble using a transition matrix
results in worse performance over the straightforward TMMC.

@@

Key words: Monte Carlo methods, Square-well fluid, Histogram methods

@@

\underline{\degree} project of \underline{\author} presented on
\underline{\date}.

@@

\hrule ~\\
David Roundy, Mentor, representing Physics

@@

\hrule ~\\
Henri J. F. Jansen, Committee Member, representing Physics

@@

\hrule ~\\
David H. McIntyre, Committee Member, representing Physics

@@

Algorithm \ref{alg:wang_landau} walks through the Wang-Landau
initialization procedure. This algorithm includes four free parameters
which affect initialization process: the number of iterations $n$ to
run in step \ref{alg:wang_landau_sim}, the initial factor $f_0$, the
update factor $u$, and the flatness condition $C\sp{H}$. The cutoff
$c$ determines the condition for Algorithm \ref{alg:wang_landau} to
terminate.
